---
title: "NCSU Sweetpotatobase Yield Dataset Analysis"
author: "Bryan"
date: "April 17, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Import

For this analysis I've downloaded two datasets. A spreadsheet with all the NCSU yield data available in the database (downloaded from the wizard), and a pedigree file for all the accessions with yield data (from the download page, using an accession list). Step 1 is to load the yield data into R and do some simple filtering and processing so I can look specifically the total yield per plant of accessions that are still active in the program.

```{r}

library(dplyr)
setwd("/home/bryan/BLUPsandEBVs")
yieldfile = read.csv("NCSUpheno.csv", skip = 3, header=T)
colnames(yieldfile)
data2018 <- yieldfile %>% filter(studyYear == 2018)
planted2018 <- unique(data2018$germplasmName)
yield <-  yieldfile[yieldfile$germplasmName %in% planted2018,]

yield$totalperplant <- (yield$Total.root.weight.computation.per.plot.CO_331.0000237/yield$Plants.harvested.counting.number.per.plot.CO_331.0000679)
yield <- yield[complete.cases(yield[ , which(colnames(yield)=="totalperplant")]),]

# Write and read dataset to avoid issue with filters not actually removing excess accession names. 
write.csv(yield, file ="yield.csv")
yield <- read.csv("yield.csv", header=T)

```

## Option 1: Plot Averages

Calculate the average total root yield and % of check root yield for each accession

```{r}
totalmeans <- yield %>% 
  filter(!is.na(totalperplant)) %>%
  group_by(germplasmName) %>%
  summarise(totalperplant = mean(totalperplant))
  
totalcounts <- yield %>% 
  filter(!is.na(totalperplant)) %>%
  group_by(germplasmName) %>% tally()

total <- merge(totalmeans,totalcounts,by="germplasmName") %>% arrange(desc(totalperplant))
head(total, n=15)

# Combine trial and rep as new globally unique 'block' factor
yield$block <- paste(yield$studyName, yield$replicate, sep='') 

# Extract covington's yield measurments by block and give them unique column names
cov_data <- yield[yield$germplasmName == "Covington", c("totalperplant","block")]
names(cov_data) <- c("covtotalperplant", "block")

# Merge covington values back to data frame by block, then calculate percent cov traits
yieldwithcov <- merge(yield,cov_data,by="block")
yieldwithcov$TotalYieldPerCov <- yieldwithcov$totalperplant/yieldwithcov$covtotalperplant
write.csv(yieldwithcov, file ="yield.csv")
yieldwithcov <- read.csv("yield.csv", header=T)

unique(yieldwithcov$germplasmName)

# Calculate total yield as percentage of Covington mean values
covtotalmeans <- yieldwithcov %>% 
  filter(!is.na(TotalYieldPerCov)) %>%
  group_by(germplasmName) %>%
  summarise(TotalYieldPerCov = mean(TotalYieldPerCov))
covtotalcounts <- yieldwithcov %>% 
  filter(!is.na(TotalYieldPerCov)) %>%
  group_by(germplasmName) %>% tally()

percov <- merge(covtotalmeans,covtotalcounts,by="germplasmName") %>% arrange(desc(TotalYieldPerCov))
head(percov, n=15)

```

## Option 2: BLUPs

Calculate total yield and %Cov BLUPs for each genotype, fitting year, location, rep, DAP, and their interactions in the total model, leaving them out of the % COV model

```{r}

library(lme4)

# Total Yield BLUPS
totalmodel = lmer(totalperplant~ (1|germplasmName) + locationName, data = yield, control=lmerControl(check.nlev.gtr.1="warning"))

totalintercept <- fixef(totalmodel)[1]
totalblup = ranef(totalmodel)$germplasmName + totalintercept
names(totalblup) <- c("BLUP")
totalblup$germplasmName <- rownames(totalblup)
totalblup <- totalblup[order(-totalblup$BLUP),,drop = FALSE]
head(totalblup, n=15)

# Total %COV BLUPS
covtotalmodel = lmer(TotalYieldPerCov~ (1|germplasmName), data = yieldwithcov, control=lmerControl(check.nlev.gtr.1="warning"))

covtotalintercept <- fixef(covtotalmodel)
covtotalblup = ranef(covtotalmodel)$germplasmName + covtotalintercept
names(covtotalblup) <- c("BLUP")
covtotalblup$germplasmName <- rownames(covtotalblup)
percovblup <- covtotalblup[order(-covtotalblup$BLUP),,drop = FALSE]
head(percovblup, n=15)
 
```

# 3 For more flexible analysis, define a custom mixed model function for calculating BLUPs that allows pedigree input in form of an additive relationship matrix. Based on material from Cornell's PLBRG 4080 course.

Once defined, optionally show that the same BLUPS calculated in the previous step can be produced with the custom function.

```{r}
# Custom mixed model function 

customMM <- function(alpha, y, X, Z, A) {
  # return a new 2 column data frame containing random effect estimates. These will be breeding values / EBVs if A matrix is the inverse of the additive relationship matrix, or simple BLUPs if it is a simple identity matrix.
  c11= t(X)%*%X
  c12= t(X)%*%Z
  c21= t(Z)%*%X
  c22= t(Z)%*%Z + A*alpha
  RHS= rbind(t(X)%*%y,t(Z)%*%y)
  upper_LHS= cbind(c11, c12)
  lower_LHS= cbind(c21, c22)
  LHS= rbind(upper_LHS, lower_LHS)
  invLHS = solve(LHS)
  hatvec = invLHS %*% RHS
  uhat = hatvec[-(1:ncol(X))]
  result = as.data.frame(uhat)
  result$id = colnames(Z)
  return(result %>% arrange(desc(uhat)))
}

# Total BLUPS from customMM
# totalmodel = lmer(totalperplant~ (1|germplasmName) + locationName, data = yield, control=lmerControl(check.nlev.gtr.1="warning"))
# vt = as.data.frame(VarCorr(totalmodel))
# Vg = vt$vcov[vt$grp == "germplasmName"]
# Ve = vt$vcov[vt$grp == "Residual"]
# alpha = Ve/Vg
# 
# y <- as.vector(yield$totalperplant)
# X <- model.matrix(totalperplant~locationName, data=yield)
# Z <- model.matrix(totalperplant~germplasmName-1, data=yield)
# colnames(Z)= gsub("germplasmName", "", colnames(Z))
# Ident = diag(length(colnames(Z)))
# 
# totalblup = customMM(alpha, y, X, Z, Ident)
# head(totalblup, n=15)
# 
# # Total %Cov from customMM
# covmodel = lmer(TotalYieldPerCov~ (1|germplasmName) + locationName, data = yieldwithcov, control=lmerControl(check.nlev.gtr.1="warning"))
# vt = as.data.frame(VarCorr(covmodel))
# Vg = vt$vcov[vt$grp == "germplasmName"]
# Ve = vt$vcov[vt$grp == "Residual"]
# covalpha = Ve/Vg
# 
# covy <- as.vector(yieldwithcov$TotalYieldPerCov)
# covX <- model.matrix(TotalYieldPerCov~locationName, data=yieldwithcov)
# covZ <- model.matrix(TotalYieldPerCov~germplasmName-1, data=yieldwithcov)
# colnames(Z)= gsub("germplasmName", "", colnames(Z))
# covIdent = diag(length(colnames(Z)))
# percovblup = customMM(covalpha, covy, covX, covZ, covIdent)
# head(percovblup, n=15)

```

## Option 3: EBVs

Now loading pedigree info to create an additive relationship matrix and use it to calculate EBVs

```{r}
library(pedigreemm)

#read in pedigree file download from spbase and remove duplicated rows
ped <- read.table ("pedigree.txt", header = TRUE, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)

# use editPed to order the pedigree by generation and make sure there is a row for each unique accession name, including parents
P2= editPed(dam=ped$Female_Parent, sire=ped$Male_Parent, label=ped$Accession)

# read records corrected by editPed into a formal pedigree object
P3= pedigree(P2$sire, P2$dam, P2$label)
Amat <- getA(P3)

# Retrieve the variance components from the total yield model
vt = as.data.frame(VarCorr(totalmodel))
Vg = vt$vcov[vt$grp == "germplasmName"]
Ve = vt$vcov[vt$grp == "Residual"]
alpha = Ve/Vg

y <- as.vector(yield$totalperplant)
X <- model.matrix(totalperplant~locationName, data=yield)
Z <- model.matrix(totalperplant~germplasmName-1, data=yield)
colnames(Z)= gsub("germplasmName", "", colnames(Z))

Amat= Amat[colnames(Z),colnames(Z)]
Amatinv= solve(Amat)
totalebv = customMM(alpha, y, X, Z, Amatinv)
totalebv$uhat <- totalebv$uhat + totalintercept 
head(totalebv, n=15)

# Total %Cov EBVS

covvt = as.data.frame(VarCorr(covtotalmodel))
covVg = covvt$vcov[covvt$grp == "germplasmName"]
covVe = covvt$vcov[covvt$grp == "Residual"]
covalpha = covVe/covVg

covy <- as.vector(yieldwithcov$TotalYieldPerCov)
covX <- model.matrix(TotalYieldPerCov~locationName, data=yieldwithcov)
covZ <- model.matrix(TotalYieldPerCov~germplasmName-1, data=yieldwithcov)
colnames(covZ)= gsub("germplasmName", "", colnames(covZ))
covAmat= Amat[colnames(covZ),colnames(covZ)]
covAmatinv= solve(covAmat)
percovebv = customMM(covalpha, covy, covX, covZ, covAmatinv)
percovebv$uhat <- percovebv$uhat + covtotalintercept 
head(percovebv, n=15)

```

#Compare Rankings of top 15 clones for all methods for each trait

```{r}
library(gridExtra)
library(grid)

# cbind all ranked lists and draw a table
total$"total" <- apply( total[ , c("germplasmName","totalperplant") ] , 1 , paste , collapse = " - " )
percov$"percov" <- apply( percov[ , c("germplasmName","TotalYieldPerCov") ] , 1 , paste , collapse = " - " )
totalblup$"totalBLUP" <- apply( totalblup[ , c("germplasmName","BLUP") ] , 1 , paste , collapse = " - " )
percovblup$"percovBLUP" <- apply( percovblup[ , c("germplasmName","BLUP") ] , 1 , paste , collapse = " - " )
totalebv$"totalEBV" <- apply( totalebv[ , c("id","uhat") ] , 1 , paste , collapse = " - " )
percovebv$"percovEBV" <- apply( percovebv[ , c("id","uhat") ] , 1 , paste , collapse = " - " )

totald <- as.data.frame(cbind(
  head(total$`total`, n=15), 
  head(totalblup$`totalBLUP`, n=15), 
  head(totalebv$`totalEBV`, n=15)
))

names(totald) <- c("TotalKg/plant", "TotalKg/plant BLUPs","TotalKgperplant EBVs")
grid.table(totald)

covtotald <- as.data.frame(cbind(
  head(percov$`percov`, n=15), 
  head(percovblup$`percovBLUP`, n=15),
  head(percovebv$`percovEBV`, n=15)
))

names(covtotald) <- c("TotalKg%Covington", "TotalKg%Covington BLUPs", "TotalKg%Covington EBVs")
grid.table(covtotald)

```

