---
title: "NCSU Sweetpotatobase Yield Dataset Analysis"
author: "Bryan"
date: "April 17, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Import

For this analysis I've downloaded two datasets. A spreadsheet with all the NCSU yield data available in the database (downloaded from the wizard), and a pedigree file for all the accessions with yield data (from the download page, using an accession list). The first thing I want to do is load the yield data into R and do some simple filtering and processing so I can look specifically the total yield per plant of accessions that are still active in the program.

```{r}

library(dplyr)
setwd("/home/bryan/BLUPsandEBVs")
yieldfile = read.csv("NCSUpheno.csv", skip = 3, header=T)
colnames(yieldfile)
data2018 <- yieldfile %>% filter(studyYear == 2018)
planted2018 <- unique(data2018$germplasmName)
yield <-  yieldfile[yieldfile$germplasmName %in% planted2018,]

yield$totalperplant <- (yield$Total.root.weight.computation.per.plot.CO_331.0000237/yield$Plants.harvested.counting.number.per.plot.CO_331.0000679)
yield <- yield[complete.cases(yield[ , which(colnames(yield)=="totalperplant")]),]

# Write and read dataset to avoid issue with filters not actually removing excess accession names. 
write.csv(yield, file ="yield.csv")
yield <- read.csv("yield.csv", header=T)

```

## Option 1: Plot Averages

Calculate the average total root yield and % of check root yield for each accession

```{r}
totalmeans <- yield %>% 
  filter(!is.na(totalperplant)) %>%
  group_by(germplasmName) %>%
  summarise(totalperplant = mean(totalperplant))
  
totalcounts <- yield %>% 
  filter(!is.na(totalperplant)) %>%
  group_by(germplasmName) %>% tally()

total <- merge(totalmeans,totalcounts,by="germplasmName") %>% arrange(desc(totalperplant))
head(total, n=15)

# Combine trial and rep as new globally unique 'block' factor
yield$block <- paste(yield$studyName, yield$replicate, sep='') 

# Extract covington's yield measurments by block and give them unique column names
cov_data <- yield[yield$germplasmName == "Covington", c("totalperplant","block")]
names(cov_data) <- c("covtotalperplant", "block")

# Merge covington values back to data frame by block, then calculate percent cov traits
yieldwithcov <- merge(yield,cov_data,by="block")
yieldwithcov$TotalYieldPerCov <- yieldwithcov$totalperplant/yieldwithcov$covtotalperplant
write.csv(yieldwithcov, file ="yield.csv")
yieldwithcov <- read.csv("yield.csv", header=T)

unique(yieldwithcov$germplasmName)

# Calculate total yield as percentage of Covington mean values
covtotalmeans <- yieldwithcov %>% 
  filter(!is.na(TotalYieldPerCov)) %>%
  group_by(germplasmName) %>%
  summarise(TotalYieldPerCov = mean(TotalYieldPerCov))
covtotalcounts <- yieldwithcov %>% 
  filter(!is.na(TotalYieldPerCov)) %>%
  group_by(germplasmName) %>% tally()

percov <- merge(covtotalmeans,covtotalcounts,by="germplasmName") %>% arrange(desc(TotalYieldPerCov))
head(percov, n=15)

```

## Option 2: BLUPs

Calculate total yield and %Cov BLUPs for each genotype, fitting year, location, rep, DAP, and their interactions in the total model, leaving them out of the % COV model

```{r}

# library(lme4)
# 
# # Total Yield BLUPS
# totalmodel = lmer(totalperplant~ (1|germplasmName) + locationName, data = yield, control=lmerControl(check.nlev.gtr.1="warning"))
# 
# totalblup = ranef(totalmodel)$germplasmName
# names(totalblup) <- c("BLUP")
# totalblup <- totalblup[order(-totalblup$BLUP),,drop = FALSE]
# head(totalblup, n=15)
# 
# # Total %COV BLUPS
# covtotalmodel = lmer(TotalYieldPerCov~ (1|germplasmName), data = yield, control=lmerControl(check.nlev.gtr.1="warning"))
# 
# covtotalblup = ranef(covtotalmodel)$germplasmName
# names(covtotalblup) <- c("BLUP")
# percovblup <- covtotalblup[order(-covtotalblup$BLUP),,drop = FALSE]
# head(percovblup, n=15)

```

# 2.5 As proof of concept, calculate manual BLUPS and compare to previous BLUPs

```{r}

# Retrieve the variance components from the total yield model
totalmodel = lmer(totalperplant~ (1|germplasmName) + locationName, data = yield, control=lmerControl(check.nlev.gtr.1="warning"))
vt = as.data.frame(VarCorr(totalmodel))
Vg = vt$vcov[vt$grp == "germplasmName"]
Ve = vt$vcov[vt$grp == "Residual"]
alpha = Ve/Vg

y <- as.vector(yield$totalperplant)
X <- model.matrix(totalperplant~locationName, data=yield)
Z <- model.matrix(totalperplant~germplasmName-1, data=yield)
colnames(Z)= gsub("germplasmName", "", colnames(Z))
Ident = diag(length(colnames(Z)))

c11= t(X)%*%X
c12= t(X)%*%Z
c21= t(Z)%*%X
c22= t(Z)%*%Z + Ident*alpha
RHS= rbind(t(X)%*%y,t(Z)%*%y)
upper_LHS= cbind(c11, c12)
lower_LHS= cbind(c21, c22)
LHS= rbind(upper_LHS, lower_LHS)
invLHS = solve(LHS)
hatvec = invLHS %*% RHS
betahat = hatvec[1:ncol(X)]
head(betahat)
uhat = hatvec[-(1:ncol(X))] # EBVs
BLUPs = as.data.frame(uhat)
BLUPs$germplasmName = colnames(Z)
totalblup = BLUPs %>% arrange(desc(uhat))
head(totalblup, n=15)

# Total %Cov maunal BLUPS

covmodel = lmer(TotalYieldPerCov~ (1|germplasmName) + locationName, data = yieldwithcov, control=lmerControl(check.nlev.gtr.1="warning"))
vt = as.data.frame(VarCorr(covmodel))
Vg = vt$vcov[vt$grp == "germplasmName"]
Ve = vt$vcov[vt$grp == "Residual"]
alpha = Ve/Vg

y <- as.vector(yieldwithcov$TotalYieldPerCov)
X <- model.matrix(TotalYieldPerCov~locationName, data=yieldwithcov)
Z <- model.matrix(TotalYieldPerCov~germplasmName-1, data=yieldwithcov)
colnames(Z)= gsub("germplasmName", "", colnames(Z))
Ident = diag(length(colnames(Z)))
c11= t(X)%*%X
c12= t(X)%*%Z
c21= t(Z)%*%X
c22= t(Z)%*%Z + Ident*alpha
RHS= rbind(t(X)%*%y,t(Z)%*%y)
upper_LHS= cbind(c11, c12)
lower_LHS= cbind(c21, c22)
LHS= rbind(upper_LHS, lower_LHS)
invLHS = solve(LHS)
hatvec = invLHS %*% RHS
betahat = hatvec[1:ncol(X)]
head(betahat)
uhat = hatvec[-(1:ncol(X))] # EBVs
BLUPs = as.data.frame(uhat)
BLUPs$germplasmName = colnames(Z)
percovblup = BLUPs %>% arrange(desc(uhat))
head(percovblup, n=15)

```

## Option 3: EBVs

Calculate then estimated breeding values. Essentially the same as BLUPs, but now taking into account pedigree information. I have not had much luck using pedigreemm, so instead will calcualte the EBVs manually as covered in PLBRG 4080

```{r}

# Retrieve the variance components from the total yield model
vt = as.data.frame(VarCorr(totalmodel))
Vg = vt$vcov[vt$grp == "germplasmName"]
Ve = vt$vcov[vt$grp == "Residual"]
alpha = Ve/Vg

#read in pedigree file download from spbase and remove duplicated rows
ped <- read.table ("pedigree.txt", header = TRUE, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)

# use editPed to order the pedigree by generation and make sure there is a row for each unique accession name, including parents
P2= editPed(dam=ped$Female_Parent, sire=ped$Male_Parent, label=ped$Accession)

# read records corrected by editPed into a formal pedigree object
P3= pedigree(P2$sire, P2$dam, P2$label)
Amat <- getA(P3)
dim(Amat)
#include_list <- yield$germplasmName
#Amat <- Amat[rownames(Amat) %in% include_list, colnames(Amat) %in% include_list] 

#unique(yield$germplasmName)

y <- as.vector(yield$totalperplant)
X <- model.matrix(totalperplant~locationName, data=yield)
Z <- model.matrix(totalperplant~germplasmName-1, data=yield)
colnames(Z)= gsub("germplasmName", "", colnames(Z))

Amat= Amat[colnames(Z),colnames(Z)]
Amatinv= solve(Amat)

c11= t(X)%*%X
c12= t(X)%*%Z
c21= t(Z)%*%X
c22= t(Z)%*%Z + Amatinv*alpha
RHS= rbind(t(X)%*%y,t(Z)%*%y)
upper_LHS= cbind(c11, c12)
lower_LHS= cbind(c21, c22)
LHS= rbind(upper_LHS, lower_LHS)
invLHS = solve(LHS)
hatvec = invLHS %*% RHS
betahat = hatvec[1:ncol(X)]
head(betahat)
uhat = hatvec[-(1:ncol(X))] # EBVs
EBVs = as.data.frame(uhat)
EBVs$germplasmName = colnames(Z)
totalebv = EBVs %>% arrange(desc(uhat))
head(totalebv, n=15)

# Total %Cov EBVS

vt = as.data.frame(VarCorr(covmodel))
Vg = vt$vcov[vt$grp == "germplasmName"]
Ve = vt$vcov[vt$grp == "Residual"]
alpha = Ve/Vg

y <- as.vector(yieldwithcov$TotalYieldPerCov)
X <- model.matrix(TotalYieldPerCov~locationName, data=yieldwithcov)
Z <- model.matrix(TotalYieldPerCov~germplasmName-1, data=yieldwithcov)
colnames(Z)= gsub("germplasmName", "", colnames(Z))
Amat= Amat[colnames(Z),colnames(Z)]
Amatinv= solve(Amat)
c11= t(X)%*%X
c12= t(X)%*%Z
c21= t(Z)%*%X
c22= t(Z)%*%Z + Amatinv*alpha
RHS= rbind(t(X)%*%y,t(Z)%*%y)
upper_LHS= cbind(c11, c12)
lower_LHS= cbind(c21, c22)
LHS= rbind(upper_LHS, lower_LHS)
invLHS = solve(LHS)
hatvec = invLHS %*% RHS
betahat = hatvec[1:ncol(X)]
head(betahat)
uhat = hatvec[-(1:ncol(X))] # EBVs
EBVs = as.data.frame(uhat)
EBVs$germplasmName = colnames(Z)
percovebv = EBVs %>% arrange(desc(uhat))
head(percovebv, n=15)

```

#Compare Rankings of top 15 clones for all methods for each trait

```{r}
library(gridExtra)
library(grid)

# cbind all ranked lists and draw a table
total$"total" <- apply( total[ , c("germplasmName","totalperplant") ] , 1 , paste , collapse = " - " )
percov$"percov" <- apply( percov[ , c("germplasmName","TotalYieldPerCov") ] , 1 , paste , collapse = " - " )
#totalblup$names <- rownames(totalblup)
totalblup$"totalBLUP" <- apply( totalblup[ , c("germplasmName","uhat") ] , 1 , paste , collapse = " - " )
#percovblup$names <- rownames(percovblup)
percovblup$"percovBLUP" <- apply( percovblup[ , c("germplasmName","uhat") ] , 1 , paste , collapse = " - " )
#totalebv$names <- rownames(totalebv)
totalebv$"totalEBV" <- apply( totalebv[ , c("germplasmName","uhat") ] , 1 , paste , collapse = " - " )
#percovebv$names <- rownames(percovebv)
percovebv$"percovEBV" <- apply( percovebv[ , c("germplasmName","uhat") ] , 1 , paste , collapse = " - " )

totald <- as.data.frame(cbind(
  head(total$`total`, n=15), 
  head(percov$`percov`, n=15), 
  head(totalblup$`totalBLUP`, n=15), 
  head(percovblup$`percovBLUP`, n=15),
  head(totalebv$`totalEBV`, n=15), 
  head(percovebv$`percovEBV`, n=15)
))

names(totald) <- c("TotalKg/plant", "TotalKg%Covington", "TotalKg/plant BLUPs", "TotalKg%Covington BLUPs","TotalKgperplant EBVs","TotalKg%Covington EBVs")
grid.table(totald)

```

